Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan

### 2025-06-02 analysis/paper/paper.qmd
editing

The faunal assemblage from the sampling area of Kiwulan consists of `r total_NISP` bone fragments, of which `r family_ident` were identified to at least the family level (@tbl-counts). Among those bones identified to the family or more specific level, Cervidae is the majority (`r cer_NISP_per`%), followed by Suidae (`r suid_NISP_per`%) and Bovinae (`r buff_NISP_per`%). To further compare general patterns of animal consumption based on their economic values and meat return rates, animals of similar body size and from the same family were grouped together. For example, the "deer" category includes only medium and large deer species, while the smaller deer species, muntjac, which implies a lower meat yield, was excluded and treated as a separate category. Similarly, water buffalo and yellow cattle, which are comparable in size, were grouped into the "bovine" category, and domestic pigs (CL4: n = 2; CL6: n = 1) were included in the "suid" category. Rats (*Rattus sp.*) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes. In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 (@tbl-counts, @fig-NISP-proportion: A). A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).

@@

```{r}
#| label: fig-NISP-proportion
#| echo: false
#| fig-cap: "A: The number of identified specimens per cultural layer. NISP data is derived from Table 1 . CL6 represents the earliest layer, and CL1 is the youngest. B: The weight of unidentified small bone fragments per cultural layer. C: Relative abundance of vertebrate classes by cultural layer based on normed NISP. D: Relative abundance of identified mammals by cultural layer based on NISP."

@@

<!-- population dynamics and arguments-->
The high population during the European period, as inferred by the frequencies of pottery, may have contributed to the overall increase in animal use [@Wang2020pot; @Hsieh2009]. However, the impact of foreign presence can be still be identified in deer butchery patterns, the relative proportions of animals, and animal ubiquity. First, lower limbs with skinning cuts were mostly found in the European period, followed by the late pre-European phase, aligning with the rise of regional trade in the 14th–15th centuries and its intensification in the 17th century. Second, if the high frequency of deer heads during the European period was solely due to food stress from population growth, leading to increased brain or tongue consumption, a parallel rise in marrow extraction would be expected. However, long bones with fracture patterns showing 10–30% less frequent during the European and late pre-European periods than in other periods. Third, population growth would have likely led to a more proportional increase in both large and small animals. Despite high taxonomic diversity during the European period, suid remains showed a disproportionate rise in ubiquity, quantity, and proportion. This indicates that the trade of deerskin and meat may have led Indigenous people to rely more on suids as an alternative prey due to their comparable body size and meat return rates to deer [@Broughton2011prey].

>>>>

The faunal assemblage from the sampling area of Kiwulan consists of `r total_NISP` bone fragments, of which `r family_ident` were identified to at least the family level (@tbl-counts). Among those bones identified to the family or more specific level, Cervidae is the majority (`r cer_NISP_per`%), followed by Suidae (`r suid_NISP_per`%) and Bovinae (`r buff_NISP_per`%). Based on the assumption that differences in prey body size may be associated with distinct economic values [@Broughton2011prey; @Ugan2005], animals of similar body size and from the same family were grouped together to further compare general patterns of animal use. For example, the "deer" category includes only medium and large deer species, while the small deer species, muntjac, which implies a lower ranking resource, was excluded and treated as a separate category. Similarly, water buffalo and yellow cattle, which are comparable in size, were grouped into the "bovine" category, and domestic pigs (CL4: n = 2; CL6: n = 1) were included in the "suid" category. Rats (*Rattus sp.*) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes. In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 (@tbl-counts, @fig-NISP-proportion: A). A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B), which were too fragmented to be identified even to the class level.

@@

```{r}
#| label: fig-NISP-proportion
#| echo: false
#| fig-cap: "A: The number of identified specimens per cultural layer. NISP data is derived from Table 1 . CL6 represents the earliest layer, and CL1 is the youngest. B: The weight of unidentified small bone fragments that could not be identified to the class level, shown per cultural layer. C: Relative abundance of vertebrate classes by cultural layer based on normed NISP. D: Relative abundance of identified mammals by cultural layer based on NISP."

@@

<!-- population dynamics and arguments-->
The high population during the European period, as inferred by the frequencies of pottery, may have contributed to the overall increase in animal use [@Wang2020pot; @Hsieh2009]. However, the impact of foreign presence can be still be identified in deer butchery patterns, the relative proportions of animals, and animal ubiquity. First, lower limbs with skinning cuts were mostly found in the European period, followed by the late pre-European phase, aligning with the rise of regional trade in the 14th–15th centuries and its intensification in the 17th century. Second, if the high frequency of deer heads during the European period was solely due to food stress from population growth, leading to increased brain or tongue consumption, a parallel rise in marrow extraction would be expected. However, long bones with fracture patterns showing 10–30% less frequent during the European and late pre-European periods than in other periods. Third, population growth would have likely led to a more proportional increase in both large and small animals. Despite high taxonomic diversity during the European period, suid remains showed a disproportionate rise in ubiquity, quantity, and proportion. This indicates that the trade of deerskin and meat may have led Indigenous people to rely more on suids as an alternative prey due to their comparable body size and similar return rates to deer [@Broughton2011prey].
